Should geophysicists use the gravity disturbance or the anomaly?

### 2018-05-18 manuscript/letter.tex
Remove the mentions of the purpose of reductions
Modifications made after our live discussion.

There is no reason to get into that in this paper. We're already
polemical enough.
Changed the title back to a question.
Part about free-air is a discussion of it being an approximation to the
disturbance so moved it back to the previous section.
Bouguer is now a section on the topographic masses.

\title[Gravity anomaly or gravity disturbance?]
      {Geophysicists should use the gravity disturbance rather than the gravity anomaly}

@@

\begin{summary}
 The gravity anomaly is defined as the difference between the Earth's gravity
 on the geoid and the gravity of the normal Earth on the reference ellipsoid.
 Because these quantities are not at the same point, the anomaly contains
 centrifugal accelerations and cannot be considered a harmonic function.
 The gravity disturbance is the difference between gravity and normal gravity
 at the same point.
 Consequently, the centrifugal effects can be neglected and the disturbance can
 be considered a harmonic function.
 This is the premise behind most potential-field data processing techniques
 (e.g., upward/downward continuation and vertical derivatives).
 Unlike the anomaly, the disturbance is due solely to the
 gravitational effects of geologic sources, making it the most appropriate
 choice for modelling.
 Use of the gravity anomaly in geophysics carries with it the implicit
 assumption that it is a good approximation for the gravity disturbance.
 However, bear in mind that the difference between the gravity disturbance and
 the free-air anomaly can be larger than 10 mGal worldwide.
 Fortunately, the solution to these problems is to simply replace
 heights referenced to the geoid (orthometric) with heights referenced to the
 ellipsoid (geometric) for geophysical use cases.
\end{summary}

@@

\section{The geophysical meaning of gravity reductions}

Different approximations for the gravity anomaly can be calculated
depending on the corrections applied to them.
These corrections are usually called \textit{gravity reductions}
and their aim is to approximate the magnitude of gravity at the geoid
$g_{Q^\prime}$ from the magnitude of gravity $g_P$ measured at a point $P$
on or above the surface of the Earth.

The \textit{free-air anomaly}, for example, can be defined as
\citep{blakely1996, hofmann-wellenhof-moritz2005}:

\begin{equation}
\Delta g^{F}
= \left(g_{P} - \frac{\partial \gamma}{\partial h} \, H_{P} \right)
- \gamma_{Q} \: ,
\label{eq:free-air-anomaly}
\end{equation}

\noindent
where $\partial \gamma/\partial h \approx -0.3086$ mGal/m is the
derivative of the normal gravity with respect to the geometric height $h$
and $H_{P}$ is the orthometric height (i.e., with respect to the geoid)
of point $P$ (Fig. \ref{fig:surfaces}).
The term between parenthesis represents an
approximation of the gravity at the point $Q^{\prime}$ on the geoid,
disregarding the effect of topographic masses.
Similarly, we can apply the same correction
to calculate the gravity disturbance by
approximating the normal gravity $\gamma_{P}$
at a point $P$ on the Earth's surface
from the normal gravity at the ellipsoid $\gamma_{Q}$:

@@

\noindent
where $h_P$ is the geometric height of point P.
Such an approximation is usually not required because
an analytical expression exists for calculating $\gamma$ at any point
on or above the ellipsoid \citep{li2001}.
The same cannot be said for the free-air anomaly because
there is usually no way of measuring gravity at the geoid ($g_{Q^\prime}$).

@@

The \textit{Bouguer anomaly}
has the aim of removing the gravitational effect of the topographic masses
between the observation point $P$ and the point on the geoid $Q^\prime$.
It is defined in the continents as
\citep{blakely1996, hofmann-wellenhof-moritz2005}:

\begin{equation}
\Delta g^{B} =
\left(g_{P}
    - \frac{\partial \gamma}{\partial h} \, H_{P}
    - 2 \pi G \rho_{t} \, H_{P} \right)
- \gamma_{Q} \: ,
\label{eq:simple-bouguer-anomaly}
\end{equation}

\noindent
where $\rho_t$ is the density, $G$ is the gravitational constant,
and the rightmost term in parenthesis is the simple Bouguer correction.
This term approximates the gravitational effect that all topographic
masses above the geoid exert on point $P$ by the effect of a homogeneous,
infinitely extended slab of constant density $\rho_{t}$ and thickness $H_{P}$.
The same correction can be applied to the gravity disturbance
(using the approximation in eq. \ref{eq:gravity-disturbance-approx})
with the aim of removing the effect of topographic masses between
the observation point $P$ and a point on the ellipsoid $Q$.
We will call this quantity the \textit{Bouguer disturbance}:

@@

At first glance, the only difference between eqs.
\ref{eq:simple-bouguer-anomaly} and \ref{eq:bouguer-disturbance} is the use of
orthometric ($H_P$) versus geometric ($h_P$) heights.
However, their objectives and geophysical interpretations are significantly
different.
The corrections applied in eq. \ref{eq:simple-bouguer-anomaly} approximate
gravity at the geoid from measurements made on the Earth's surface by removing
the effects of the distance and the topographic masses between the two points.
Apart from the centrifugal effect mentioned in the previous section,
the gravitational effect of topographic masses between the geoid and ellipsoid
are still present in the anomaly.
Conversely, when calculating the Bouguer disturbance one removes from
the observed gravity the effects of an ellipsoidal reference model (the normal
Earth) and the known topographic masses between the surface of the ellipsoid
and observation point.
Hence, the Bouguer disturbance can be interpreted as being solely the
gravitational effect produced by the density contrasts between the actual
internal density distribution of the Earth and that of the normal Earth.

There was a certain lack of comprehension regarding the geophysical meaning of
gravity anomalies until the mid 90s.
As properly pointed out by \citet{chapin1996},
``although the corrections which bring about a Bouguer gravity anomaly are well
established, the reasons for doing them are not well understood. One cause of
this common misunderstanding is that the subject has been poorly presented in
many of the basic texts''.
In his seminal book, \citet{blakely1996} shed some light on the geophysical
meaning of gravity anomalies from the perspective of applied geophysics and
their relationship to the gravity sources.
We complement this interpretation by stressing the fact that
gravity anomalies reflect not only the gravity sources,
but also a combination of gravitational and centrifugal effects.
These additional, non-harmonic and undesired effects are due to the calculation
of the normal gravity at a point other than that were gravity is measured.
In practice, this issue can be avoided entirely by performing the free-air and
Bouguer corrections using geometric instead of orthometric heights.

@@

Fortunately, all of these issues are nullified during data processing by
substituting orthometric heights (i.e., with respect to the geoid) with
geometric heights (i.e., with respect to the ellipsoid).
For modern acquisitions, this requires no extra work because GPS already
provide geometric heights.
The free-air correction can even be bypassed entirely.
As a bonus, the gravity disturbance is conceptually simpler to understand and
doesn't require the concept of the geoid with which learners often struggle.

>>>>

\title[Gravity anomaly or gravity disturbance?]
      {Should geophysicists use the gravity disturbance or the anomaly?}

@@

\begin{summary}
 The gravity anomaly is defined as the difference between the Earth's gravity
 on the geoid and the gravity of the normal Earth on the reference ellipsoid.
 Because these quantities are not at the same point, the anomaly contains
 centrifugal accelerations and cannot be considered a harmonic function.
 The gravity disturbance is the difference between gravity and normal gravity
 at the same point.
 Consequently, the centrifugal effects can be neglected and the disturbance can
 be considered a harmonic function.
 This is the premise behind most potential-field data processing techniques
 (e.g., upward/downward continuation and vertical derivatives).
 Unlike the anomaly, the disturbance is due solely to the
 gravitational effects of geologic sources, making it the most appropriate
 for geophysical purposes.
 Use of the gravity anomaly in geophysics carries with it the implicit
 assumption that it is a good approximation for the gravity disturbance.
 However, bear in mind that the difference between the gravity disturbance and
 the free-air anomaly can be larger than 10 mGal worldwide.
 SOMETHING ABOUT MODELLING.
\end{summary}

@@

Different approximations for the gravity anomaly can be calculated
depending on the corrections applied to them.
These corrections are usually called \textit{gravity reductions}.
The \textit{free-air anomaly}, for example, can be defined as
\citep{blakely1996, hofmann-wellenhof-moritz2005}:

\begin{equation}
\Delta g^{F} = g_{P}
- \gamma_{Q} - \frac{\partial \gamma}{\partial h} \, H_{P} \: ,
\label{eq:free-air-anomaly}
\end{equation}

\noindent
where the rightmost term is called the \textit{free-air correction},
$\partial \gamma/\partial h \approx -0.3086$ mGal/m is the
derivative of the normal gravity with respect to the geometric height $h$,
and $H_{P}$ is the orthometric height (i.e., with respect to the geoid)
of point $P$ (Fig. \ref{fig:surfaces}).
The free-air correction can be interpreted as a downward continuation of
$g_P$ to the geoid or as an upward continuation of $\gamma_Q$ to a height
$H_P$.
Regardless, we can also apply the free-air correction
to calculate the gravity disturbance by
approximating the normal gravity $\gamma_{P}$
at a point $P$ on the Earth's surface
from the normal gravity at the ellipsoid $\gamma_{Q}$:

@@

\noindent
where $h_P$ is the geometric height of point P.
However, using the free-air correction to calculate the disturbance is
discouraged because an analytical expression exists for calculating $\gamma$ at
any point on or above the ellipsoid \citep{li2001}.

@@

\section{The effect of topographic masses}

The \textit{Bouguer correction} can be applied to the free-air anomaly
to remove the gravitational effect of the topographic masses
between the observation point $P$ and the point on the geoid $Q^\prime$.
The \textit{Bouguer anomaly} is defined in the continents as
\citep{blakely1996, hofmann-wellenhof-moritz2005}:

\begin{equation}
\Delta g^{B} = g_{P}
    - \gamma_{Q}
    - \frac{\partial \gamma}{\partial h} \, H_{P}
    - 2 \pi G \rho_{t} \, H_{P}
     \: ,
\label{eq:simple-bouguer-anomaly}
\end{equation}

\noindent
where $\rho_t$ is the density of the topographic masses and $G$ is the
gravitational constant.
Th Bouguer correction approximates the gravitational effect that all
topographic masses above the geoid exert on point $P$ by the effect of a
homogeneous, infinitely extended slab of constant density $\rho_{t}$ and
thickness $H_{P}$.
A similar correction can be applied to the gravity disturbance
(using the approximation in eq. \ref{eq:gravity-disturbance-approx})
to remove the effect of topographic masses between
the observation point $P$ and a point on the ellipsoid $Q$.
We will call this quantity the \textit{Bouguer disturbance}:

@@

At first glance, the only difference between eqs.
\ref{eq:simple-bouguer-anomaly} and \ref{eq:bouguer-disturbance} is the use of
orthometric ($H_P$) versus geometric ($h_P$) heights.
However, their geophysical interpretations are significantly different.
Apart from the centrifugal effect mentioned in the previous section,
the Bouguer anomaly (eq. \ref{eq:simple-bouguer-anomaly}) still contains
the gravitational effect of topographic masses between the geoid and ellipsoid.
Conversely, when calculating the Bouguer disturbance one removes the effects of
an ellipsoidal reference model (the normal Earth) and the known topographic
masses between the surface of the ellipsoid and observation point.
Hence, the Bouguer disturbance can be interpreted as being solely the
gravitational effect produced by the density contrasts between the actual
internal density distribution of the Earth and that of the normal Earth plus a
homogeneous topography.

There was a certain lack of comprehension regarding the geophysical meaning of
gravity anomalies until the mid 90s.
As properly pointed out by \citet{chapin1996},
``although the corrections which bring about a Bouguer gravity anomaly are well
established, the reasons for doing them are not well understood. One cause of
this common misunderstanding is that the subject has been poorly presented in
many of the basic texts''.
In his seminal book, \citet{blakely1996} shed some light on the geophysical
meaning of gravity anomalies from the perspective of applied geophysics and
their relationship to the gravity sources.
We complement this interpretation by stressing the fact that
gravity anomalies reflect not only the gravity sources,
but also a combination of gravitational and centrifugal effects.
These additional, non-harmonic and undesired effects are due to the calculation
of the normal gravity at a point other than that were gravity is measured.

@@

In practice, most of these issues can be nullified during data processing by
substituting orthometric heights (i.e., with respect to the geoid) with
geometric heights (i.e., with respect to the ellipsoid).
For modern acquisitions, this requires no extra work because GPS already
provide geometric heights.
The free-air correction can even be bypassed entirely by computing $\gamma_P$
directly.
As a bonus, the gravity disturbance is conceptually simpler to understand and
doesn't require the concept of the geoid with which learners often struggle.
